Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-25 paper-new/appendix.tex
fix build issues, progress on CBPV semantics

>>>>

\subsection{Morphisms of CBPV Models}

There are two relevant notions of \emph{morphism} of CBPV models:
\emph{strict} and \emph{lax}.
% morphisms of CBPV models
Given call-by-push-value models
$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
that strictly presere the constructors:
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) = 1_2$
\end{enumerate}
As well as strictly preserving the corresponding universal morphisms
and coherence isomorphisms.

A lax morphism instead preserves these only up to transformation
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
\end{enumerate}
Additionally a lax morphism should have a relationship between these
transformations and the universal morphisms, but we will only consider
lax morphisms of thin categories, where such conditions hold
trivially.

\subsection{Kleisli Actions of CBPV Type Constructors}

### 2024-01-25 paper-new/intro.tex
fix build issues, progress on CBPV semantics

With this notion of abstract categorical model in hand, denotational
semantics is then the work of constructing concrete models that
exhibit the categorical construction. New and Licata
\cite{new-licata18} present such a model using categories of
$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
Maillard, Tabareau and Tanter to prove graduality of a gradual
dependently typed calculus $\textrm{CastCIC}^{\mathcalG}$. This
domain-theoretic approach meets our criteria of being a semantic
framework for proving graduality, but suffers from the limitations of
classical domain theory: the inability to model viciously
self-referential structures such as higher-order extensible state and
similar features such as runtime-extensible dynamic types. Since these
features are quite common in dynamically typed languages, we seek a
new denotational framework that can model these type system features.

>>>>

With this notion of abstract categorical model in hand, denotational
semantics is then the work of constructing concrete models that
exhibit the categorical construction. New and Licata
\cite{new-licata18} present such a model using categories of
$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
Maillard, Tabareau and Tanter to prove graduality of a gradual
dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
domain-theoretic approach meets our criteria of being a semantic
framework for proving graduality, but suffers from the limitations of
classical domain theory: the inability to model viciously
self-referential structures such as higher-order extensible state and
similar features such as runtime-extensible dynamic types. Since these
features are quite common in dynamically typed languages, we seek a
new denotational framework that can model these type system features.
